Bayesian multivariate skew-normal finite mixture model for analysis of infant development trajectories

### 2019-07-09 manuscript/manuscript.tex
sim 1

%  ALWAYS USE THE referee OPTION WITH PAPERS SUBMITTED TO BIOMETRICS!!!
%  You can see what your paper would look like typeset by removing
%  the referee option.  Because the typeset version will be in two
%  columns, however, some of your equations may be too long. DO NOT
%  use the \longequation option discussed in the user guide!!!  This option
%  is reserved ONLY for equations that are impossible to split across 
%  multiple lines; e.g., a very wide matrix.  Instead, type your equations 
%  so that they stay in one column and are split across several lines, 
%  as are almost all equations in the journal.  Use a recent version of the
%  journal as a guide. 
%  
\documentclass[useAMS,referee]{biom}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{blindtext}
\usepackage{booktabs}
\usepackage{breqn}
\usepackage{algorithm}
\usepackage{algcompatible}
\usepackage{algpseudocode}
\usepackage{float}
\usepackage{dutchcal}
\usepackage{amsfonts}

@@

Polson et al. (2013) introduce a data augmentation scheme using the P\'olya-Gamma distribution which allows for sampling of multinomial regression parameters using straightforward Gibb's updates from Gaussian full conditional distributions. In addition to more convinient parameter estimation, the P\'olya-Gamma data augmentation method for logistic regression has the advantage of direct sampling from the posterior distributions of multinomial parameters. This approach avoids the need for approximations of the posterior distribution, thus yeilding more stable sampling, especially when the number of parameters approaches the number of observations (Polson et al., 2013). P\'olya-Gamma data augmentation for multinomial regression has not yet been applied to the analysis of longitudinally clustered data.

@@

@@

A primary goal of the Nurture study is to identify clusters of infants characterized by distinct motor development trajectories throughout the first year of life. To address this aim, we propose a flexible finite mixture model that accommodates relevant features of the data, such as skewness and dependence among the responses. For $i = 1,...,n$, let $\mathbf{y}_{i}=(y_{i1},\ldots,y_{iJ})^T$ be a $J \times 1$ vector of responses (i.e., Bayley composite scores) for subject $i$ across the $J$ measurement occasions. For analysis of the Nurture data, we propose a finite mixture model of the form
\begin{equation}
	\label{eq:mixture}
f(\mathbf{y}_i) = \sum_{k = 1}^{K} \pi_{ki} f(\mathbf{y}_i|\boldsymbol\theta_k),
\end{equation}
where $\boldsymbol\theta_k$ is the set of parameters specific to cluster $k$ ($k = 1,...,K$) and $\pi_{ki}$ is a subject-specific mixing weight representing the probability that subject $i$ belongs to cluster $k$. For now we assume that $K$ is fixed; in Section \ref{s:mcmcmisc}, we discuss model selection strategies for choosing the optimal value of $K$.

To facilitate posterior inference, we introduce a latent cluster indicator variable $z_i$ taking the value $k \in \{1,...,K\}$ with probability $\pi_{ki}$. Conditional on $z_i = k$, we assume $\mathbf{y}_{i}$ is distributed as
\begin{equation}
\mathbf{y}_{i}|(z_i=k) \sim \text{MSN}_J(\boldsymbol\zeta_{ki},\boldsymbol\alpha_k,\boldsymbol\Omega_k), \label{eq:msndens}
\end{equation}
where $\text{MSN}_J(\cdot)$ denotes the $J$-dimensional multivariate skew normal density, $\boldsymbol\zeta_{ki}$ is a $J \times 1$ vector of subject- and cluster-specific location parameters, $\boldsymbol\alpha_k$ is a $J \times 1$ vector of cluster-specific skewness parameters, and $\boldsymbol\Omega_k$ is a $J \times J$ cluster-specific scale matrix that captures dependence among the $J$ responses for subject $i$. The vector $\boldsymbol\alpha_k$ has components $\alpha_{kj}$, $j = 1,...,J$, that control the skewness of outcome $j$ in cluster $k$. When $\boldsymbol\alpha_k = \mathbf{0}$, the MSN distribution reduces to the multivariate Normal distribution $\text{N}_J(\boldsymbol\zeta_{ki},\boldsymbol\Omega_k)$, where $\boldsymbol\zeta_{ki}$ is a $J \times 1$ mean vector and $\boldsymbol\Omega_k$ is a $J \times J$ covariance matrix (Azzalini and Dalla Valle, 1996).

@@

Of note, the MSN density can be expressed more compactly in terms of the Matrix Skew Normal (MatSN) density (Chen and Gupta 2005). As we will see in Section 3.6, the matrix representation of the MSN distribution admits convenient conjugate prior distributions for the regression parameters and scale matrices, which in turn leads to efficient Gibbs sampling for posterior inference. Let $\mathbf{Y}_k$ be an ${n_k \times J}$ response matrix with rows $\mathbf{y}_i^T$, $(i = 1,...,n_k)$, where $n_k = \sum_{i = 1}^n {1}_{(z_i = k)}$ is the number of observations in cluster $k$. From equation (\ref{eq:msnreg}), it follows that $\mathbf{Y}_k$ is distributed as
\begin{eqnarray}
\mathbf{Y}_{k} &\sim& \text{MatSN}_{n_k \times J}(\mathbf{M}_k,\boldsymbol\alpha_k,\mathbf{I}_{n_k},\boldsymbol\Omega_k)\label{eq:matsn}\\
%\text{vec}(\mathbf{Y}_k) =  (\mathbf{y}^T_1,...,\mathbf{y}^T_{n_k})^T\\
\text{vec}(\mathbf{M}_k) &=& (\boldsymbol\zeta_{k1}^T,...,\boldsymbol\zeta_{kn_k}^T)^T, \nonumber
\end{eqnarray}
where $\boldsymbol\zeta_{ki} = \mathbf{X}_i \boldsymbol\beta_k$ as in equation (\ref{eq:msnreg}), $\boldsymbol\alpha_k = (\alpha_{k1},...,\alpha_{kJ})^T$, $\mathbf{I}_{n_k}$ is the $n_k\times n_k$ identity matrix, and $\boldsymbol\Omega_k$ is the $J\times J$ scale matrix defined above in equation (\ref{eq:msndens}). From equation (\ref{eq:msnreg}), it follows that $\mathbf{Y}_{k}$, conditional on the $n_k\times 1$ vector of random effects $\mathbf{t}_{k} = (t_1,...,t_{n_k})^T$, is jointly distributed in matrix form as
\begin{equation}
\mathbf{Y}_k | \mathbf{t}_k \sim \text{MatNorm}_{n_k \times J}(\mathbf{M}^*_k, \mathbf{I}_{n_k}, \boldsymbol\Sigma_k),
\end{equation}
where $\text{MatNorm}_{n_k \times J}(\cdot)$ denotes a $n_k\times J$ matrix normal density, $\text{vec}(\mathbf{M}^*_k) = \mathbf{X}_{k} \boldsymbol\beta_k + \mathbf{t}_k \otimes \boldsymbol\psi_k$ is an $n_kJ\times 1$ mean vector, $\mathbf{X}_k$ is an $n_kJ \times Jp$ design matrix, $\boldsymbol\beta_k$ is the $Jp\times 1$ vector of regression coefficients defined in equation (\ref{eq:msnreg}), and $\boldsymbol\Sigma_k$ is the $J\times J$ conditional covariance of $\boldsymbol\epsilon_{ik}$ given in equation (\ref{eq:msnreg}).

\subsection{Multinomial Regression for the Cluster Indicators}
\label{s:multinom}
To accommodate heterogeneity in the cluster-membership probabilities, we model $\pi_{ki}$ as a function of covariates using a multinomial logit model 
\begin{eqnarray}
\label{eq:pik}
\pi_{ki} = \Pr(z_i = k|\mathbf{w}_i) = \frac{\text{e}^{\mathbf{w}_i^T \boldsymbol\delta_k}}{\sum_{h = 1}^K \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_{h}}},~ k=1,\ldots,K,
\end{eqnarray}
where $\mathbf{w}_i$ is an $r\times 1$ vector of subject-level covariates, $\boldsymbol\delta_k$ is a $r\times 1$ vector of regression parameters associated with membership in cluster $k$. For identifiability purposes, we fix the reference category $k = K$ and set $\boldsymbol\delta_K = \mathbf{0}$. Under this model, $z_i|\boldsymbol\pi_i \sim \text{Multinomial}(1,\boldsymbol\pi_i)$, where $\boldsymbol\pi_i = (\pi_{1i},...,\pi_{Ki})$. During MCMC estimation, the cluster labels $z_i$ are updated from their multinomial full conditional distribution and used in the remaining MCMC steps as cluster assignments as detailed later in Section \ref{s:postinf}. By allowing the cluster probabilities to vary across subjects, our model can be viewed as a \textit{mixture of experts} model in which $\pi_{ki}$ acts as a \textit{gating function} controlling the prior probability of membership in cluster $k$, and $f(\mathbf{y}_i|\boldsymbol\theta_k)$ in equation (\ref{eq:mixture}) is the ``expert'' providing information on the within-cluster distribution of $\mathbf{y}_i$ (Bishop 2016).

To facilitate sampling, we adopt an efficient data-augmentation approach introduced by Polson {et al.} (2013), which expresses the inverse-logit function as a mixture P\'olya--Gamma densities. By using P\'olya--Gamma data augmentation for the multinomial model, we obtain a \textit{P\'olya--Gamma mixture of experts model}, a computationally efficient way to obtain inferences for the mixing weights in the Bayesian setting. A random variable ${w}$ is said to follow a P\'olya--Gamma distribution with parameters $b > 0$ and $c \in \mathbb{R}$ if
\begin{equation}
{w} \sim \text{PG}(b,c) \stackrel{d}{=} \frac{1}{2\pi^2}\sum_{s=1}^{\infty}\frac{g_s}{(s-1/2)^2 + c^2/(4\pi^2)}, \label{eq:pg1}
\end{equation}
where $g_s \stackrel{iid}{\sim} \text{Ga}(b,1)$ for $s = 1,...,\infty$ . Polson {et al.} establish that for $a,\eta \in \mathbb{R}$,

@@

Polson et al. use these results to show that, in the logistic model, the Bernoulli likelihood can be written in terms of the kernel of a Normal density and a $\text{PG}(b,0)$ density, resulting in a Normal full conditional distribution for the logistic regression parameters. To extend the P\'olya--Gamma data augmentation approach to the multinomial model, we first introduce the binary indicators $U_{ki}$, such that $U_{ki} = 1_{z_i = k}$ -- i.e. $U_{ki}$ is an indicator variable taking the value $1$ if subject $i$ belongs to cluster $k$, and $0$ otherwise. The full conditional distribution of $\boldsymbol\delta_k$ is then given as  $$p(\boldsymbol\delta_k|\mathbf{U}_k,\boldsymbol\delta_{h \ne k}) \propto p(\boldsymbol\delta_k) \prod_{i = 1}^{n} \pi_{ki}^{U_{ki}}(1-\pi_{ki})^{1-U_{ki}},$$

@@

which is a logistic likelihood of the same form used by Polson et al. to derive the data-augmented P\'olya--Gamma Gibbs sampler for the logistic regression model. We can thus apply the P\'olya--Gamma sampler described in Polson et al. for logistic regression to update each $\boldsymbol\delta_k$ one at a time based on the binary indicators $U_{ki}$. To accomplish this, we first define for $k = 1,...,K$, the $n \times 1$ vector $\mathbf{U}^*_{k} = \left( \frac{U_{k1}-1/2}{w_{k1}},...,\frac{U_{kn}-1/2}{w_{kn}} \right )^T$. Polson et al. show that, conditional on $\boldsymbol{w} = (w_{k1},...,w_{kn})^T$, $\mathbf{U}^*_{k}$ follows a $\text{N}_n (\boldsymbol\eta_k,\mathbf{O}_k^{-1})$ distribution with mean $\boldsymbol\eta_k = (\eta_{k1},...,\eta_{kn})^T$ and precision matrix $\mathbf{O}_k = \text{Diag}(w_{k1},...,w_{kn})$. Thus, it follows that the full conditional distribution of $\boldsymbol\delta_k$ is given by 

@@

To accommodate intermittent missing at random (MAR) responses, we propose a convenient imputation algorithm that can be implemented ``online'' -- that is, as part of the Gibbs sampler. In Section 6, we discuss extensions to allow for non-ignorable missingness (i.e. observations missing not at random). Suppose $\mathbf{y}_i$ has $q_i\in (1,\ldots,J)$ observed values, denoted $\mathbf{y}^{obs}_i$, and $J-q_i$ intermittent missing values, denoted $\mathbf{y}^{miss}_i$. We can make use of the stochastic representation given in equation (\ref{eq:msnreg}) to impute $\mathbf{y}^{miss}_i$ from its conditional multivariate Normal distribution given $(z_i,t_i,\mathbf{y}^{obs}_i)$:

@@

The location vector $\boldsymbol\mu_{ki}$ is defined as $\boldsymbol\mu_{ki} = \mathbf{X}_i \boldsymbol\beta_k + t_i \boldsymbol\psi_k$ and is partitioned into $\boldsymbol\mu_{ki}^{miss}$ and $\boldsymbol\mu_{ki}^{obs}$ with respect to the missing and observed indices of $\mathbf{y}_i$, respectively. The partition $\boldsymbol\Sigma_{k11}$ is a $(J-q_i) \times (J-q_i)$ matrix containing the rows and columns of $\boldsymbol\Sigma_k$ corresponding to inicies of $\mathbf{y}_{i}$ where missingness occurs. Similarly, $\boldsymbol\Sigma_{k12}$ is a $(J-q_i) \times q_i$ matrix containing the rows of $\boldsymbol\Sigma_k$ that correspond to missing indices of $\mathbf{y}_i$, but columns of $\boldsymbol\Sigma_k$ that correspond to observed indices of $\mathbf{y}_i$. The remaining partitions $\boldsymbol\Sigma_{k21}, \ \text{and} \ \boldsymbol\Sigma_{k22}$ are defined in the same manner. These results follow from conventional multivariate normal theory, which we detail in the Web Appendix A. We note that, while conditional on $t_i$, $\mathbf{y}_i$ follows a MVN distribution, after marginalizing over $t_i$ the vector $\mathbf{y}_i$ follows a MSN distribution. Thus, this proposed online conditional imputation method provides a convinient way of imputing MSN responses using samples from more standard densities.

An attractive practical feature of this imputation algorithm is that it avoids multiplicative run-time scaling in $m$, the number of imputations (Gelman {et al.} 2013; Zhou and Reiter, 2010). Our approach also provides more opportunities to explore the missing data parameter space than does multiple imputation, since each missing component is drawn once per MCMC iteration, and often in practice $n_{sim} >> m$, where $n_{sim}$ is the total number of MCMC iterations (\textbf{find a reference}). In Section 4, we conduct simulation studies to demonstrate that imputing the missing MSN responses with our online conditional imputation method improves inferences over standard Bayesian multiple imputation as defined by Gelman et al. 2013.

@@

\subsubsection{Prior Specification}
We adopt a fully Bayesian inferential approach and assign prior distributions to all model parameters. Conveniently, all parameters admit conditionally conjugate priors, which greatly improves posterior computation via a data-augmented Gibbs sampler. To make use of the Matrix Normal representation introduced previously in Section 3, we define the matrix of regression parameters $\mathbb{B}^*_k$, where $\text{vec}(\mathbb{B}^*_k) = \boldsymbol\beta^*_k = (\boldsymbol\beta_k^T,\boldsymbol\psi_k^T)^T$. We assign $\mathbb{B}^*_k|\boldsymbol\Sigma_k$ a $\text{MatNorm}(\mathbf{B}^*_{0k},\mathbf{I}_{p+1},\boldsymbol\Sigma_k)$ prior, where $\mathbf{B}^*_{0k}$ is a matrix of location parameters such that $\text{vec}(\mathbf{B}^*_{0k})$ is a vector of prior location parameters for the components of $\boldsymbol\beta^*_k$, $\mathbf{I}_{p+1}$ is the $(p+1)$-dimension identity matrix, and $\boldsymbol\Sigma_k$ is the covariance matrix defined in equation (\ref{eq:msndens}), for which we specify an $\text{IW}(\mathbf{V}_{0k},\nu_{0k})$ prior. This leads to a Matrix-Normal-Inverse-Wishart joint prior for $\mathbb{B}^*_k$ and $\boldsymbol\Sigma_k$, which is the conjugate joint prior for the regression parameters in the Matrix Normal model given in equation (\ref{eq:matsn}). This induces convenient closed-form full conditional distributions that can be easily updated within our proposed Gibbs sampler.

For the multinomial logit model component, the regression parameters $\boldsymbol\delta_k = (\delta_{k1},...,\delta_{kr})$ are assigned a $\text{N}_r(\mathbf{d}_{0k}, \mathbf{S}_{0k})$ prior for $k = 1,...,K-1$, which is conditionally conjugate as described in Section \ref{s:multinom}. We allow the Normal-Inverse-Wishart and multinomial hyperparameters to vary across clusters, though they may be shared across clusters in practice. An advantage of allowing for cluster-specific prior parameters is that \textit{a priori} knowledge of motor development trends can be incorporated into certain clusters while allowing the priors for other clusters to be less informative. Additionally, prior information regarding the effect of certain covariates on cluster membership can be incorporated in to the model by choosing informative values for $\mathbf{d}_{0k} \ \text{and} \ \mathbf{S}_{0k}$

\subsubsection{Posterior Inference} 
\label{s:postinf}
The above prior specification induces closed-form full conditionals that can be efficiently updated as part of a Gibbs sampler outlined below. A programatic sketch of our MCMC algorithm is given in Web Appendix B. Additional details on derivations of full conditionals can be found in the Web Appendix A. We provide a report of MCMC diagnostics in Web Appendix C.

>>>>

%  ALWAYS USE THE referee OPTION WITH PAPERS SUBMITTED TO BIOMETRICS!!!
%  You can see what your paper would look like typeset by removing
%  the referee option.  Because the typeset version will be in two
%  columns, however, some of your equations may be too long. DO NOT
%  use the \longequation option discussed in the user guide!!!  This option
%  is reserved ONLY for equations that are impossible to split across 
%  multiple lines; e.g., a very wide matrix.  Instead, type your equations 
%  so that they stay in one column and are split across several lines, 
%  as are almost all equations in the journal.  Use a recent version of the
%  journal as a guide. 
%  
\documentclass[useAMS,referee]{biom}
\usepackage{amsmath}
\usepackage{bbm}
\usepackage{dsfont}
\usepackage{graphicx}
\usepackage{blindtext}
\usepackage{booktabs}
\usepackage{breqn}
\usepackage{algorithm}
\usepackage{algcompatible}
\usepackage{algpseudocode}
\usepackage{float}
\usepackage{dutchcal}
\usepackage{amsfonts}

@@

Polson et al. (2013) introduce a data augmentation scheme using the P\'olya-Gamma distribution which allows for sampling of multinomial regression parameters using straightforward Gibb's updates from Gaussian full conditional distributions. In addition to more convenient parameter estimation, the P\'olya-Gamma data augmentation method for logistic regression has the advantage of direct sampling from the posterior distributions of multinomial parameters. This approach avoids the need for approximations of the posterior distribution, thus yielding more stable sampling, especially when the number of parameters approaches the number of observations (Polson et al., 2013). P\'olya-Gamma data augmentation for multinomial regression has not yet been applied to the analysis of longitudinally clustered data.

@@

\begin{figure}[h]
	\includegraphics[width = 1\textwidth]{bayley_resids_plot_bytime.jpg}
\end{figure}

@@

A primary goal of the Nurture study is to identify clusters of infants characterized by distinct motor development trajectories throughout the first year of life. To address this aim, we propose a flexible finite mixture model that accommodates relevant features of the data, such as skewness, missing values, and dependence among the responses. For $i = 1,...,n$, let $\mathbf{y}_{i}=(y_{i1},\ldots,y_{iJ})^T$ be a $J \times 1$ vector of responses (i.e., Bayley composite scores) for subject $i$ across the $J$ measurement occasions. For analysis of the Nurture data, we propose a finite mixture model of the form
\begin{equation}
	\label{eq:mixture}
f(\mathbf{y}_i) = \sum_{k = 1}^{K} \pi_{ki} f(\mathbf{y}_i|\boldsymbol\theta_k),
\end{equation}
where $\boldsymbol\theta_k$ is the set of parameters specific to cluster $k$ ($k = 1,...,K$) and $\pi_{ki}$ is a subject-specific mixing weight representing the probability that subject $i$ belongs to cluster $k$. For now we assume that $K$ is fixed; in Section \ref{s:mcmcmisc}, we discuss model selection strategies for choosing the optimal value of $K$.

To facilitate posterior inference, we introduce a latent cluster indicator variable $z_i$ taking the value $k \in \{1,...,K\}$ with probability $\pi_{ki}$. Conditional on $z_i = k$, we assume $\mathbf{y}_{i}$ is distributed as
\begin{equation}
\mathbf{y}_{i}|(z_i=k) \sim \text{MSN}_J(\boldsymbol\zeta_{ki},\boldsymbol\alpha_k,\boldsymbol\Omega_k), \label{eq:msndens}
\end{equation}
where $\text{MSN}_J(\cdot)$ denotes the $J$-dimensional multivariate skew normal density, $\boldsymbol\zeta_{ki}$ is a $J \times 1$ vector of subject- and cluster-specific location parameters, $\boldsymbol\alpha_k$ is a $J \times 1$ vector of cluster-specific skewness parameters, and $\boldsymbol\Omega_k$ is a $J \times J$ cluster-specific scale matrix that captures dependence among the $J$ responses for subject $i$. The vector $\boldsymbol\alpha_k$ has components $\alpha_{kj}$, $j = 1,...,J$, that control the skewness of outcome $j$ in cluster $k$. When $\boldsymbol\alpha_k = \mathbf{0}$, the MSN distribution reduces to the multivariate normal distribution $\text{N}_J(\boldsymbol\zeta_{ki},\boldsymbol\Omega_k)$, where $\boldsymbol\zeta_{ki}$ is a $J \times 1$ mean vector and $\boldsymbol\Omega_k$ is a $J \times J$ covariance matrix.

@@

Of note, the MSN density can be expressed more compactly in terms of the matrix skew normal (MatSN) density (Chen and Gupta 2005). Let $\mathbf{Y}_k$ be an ${n_k \times J}$ response matrix with rows $\mathbf{y}_i^T$, $(i = 1,...,n_k)$, where $n_k = \sum_{i = 1}^n \mathds{1}_{(z_i = k)}$ is the number of observations in cluster $k$. From equation (\ref{eq:msnreg}), it follows that $\mathbf{Y}_k$ is distributed as
\begin{eqnarray}
\mathbf{Y}_{k} &\sim& \text{MatSN}_{n_k \times J}(\mathbf{M}_k,\boldsymbol\alpha_k,\mathbf{I}_{n_k},\boldsymbol\Omega_k)\label{eq:matsn}\\
%\text{vec}(\mathbf{Y}_k) =  (\mathbf{y}^T_1,...,\mathbf{y}^T_{n_k})^T\\
\text{vec}(\mathbf{M}_k) &=& (\boldsymbol\zeta_{k1}^T,...,\boldsymbol\zeta_{kn_k}^T)^T, \nonumber
\end{eqnarray}
where $\mathbf{M}_k$ is an $n_k \times J$ location matrix with rows $\boldsymbol\zeta_{ki} = \mathbf{X}_i \boldsymbol\beta_k$ as in equation (\ref{eq:msnreg}), $\boldsymbol\alpha_k = (\alpha_{k1},...,\alpha_{kJ})^T$, $\mathbf{I}_{n_k}$ is the $n_k\times n_k$ identity matrix, and $\boldsymbol\Omega_k$ is the $J\times J$ scale matrix defined above in equation (\ref{eq:msndens}). From equation (\ref{eq:msnreg}), it follows that $\mathbf{Y}_{k}$, conditional on the $n_k\times 1$ vector of random effects $\mathbf{t}_{k} = (t_1,...,t_{n_k})^T$, is jointly distributed in matrix form as
\begin{equation}
\mathbf{Y}_k | \mathbf{t}_k \sim \text{MatNorm}_{n_k \times J}(\mathbf{M}^*_k, \mathbf{I}_{n_k}, \boldsymbol\Sigma_k),
\end{equation}
where $\text{MatNorm}_{n_k \times J}(\cdot)$ denotes a $n_k\times J$ matrix normal density, $\mathbf{M}^*_k$ is an $n_k \times J$ matrix such that $\text{vec}(\mathbf{M}^*_k) = \mathbf{X}_{k} \boldsymbol\beta_k + \mathbf{t}_k \otimes \boldsymbol\psi_k$ is an $n_kJ\times 1$ mean vector, $\mathbf{X}_k$ is an $n_kJ \times Jp$ design matrix, $\boldsymbol\beta_k$ is the $Jp\times 1$ vector of regression coefficients defined in equation (\ref{eq:msnreg}), and $\boldsymbol\Sigma_k$ is the $J\times J$ conditional covariance of $\boldsymbol\epsilon_{ik}$ given in equation (\ref{eq:msnreg}). As described in Section 3.6, the matrix representation of the MSN distribution admits convenient conjugate prior distributions for the regression parameters and scale matrices, which in turn leads to efficient Gibbs sampling for posterior inference.

\subsection{Multinomial Regression on Cluster Indicators}
\label{s:multinom}
To accommodate heterogeneity in the cluster-membership probabilities, we model $\pi_{ki}$ as a function of covariates using a multinomial logit model 
\begin{eqnarray}
\label{eq:pik}
\pi_{ki} = \Pr(z_i = k|\mathbf{w}_i) = \frac{\text{e}^{\mathbf{w}_i^T \boldsymbol\delta_k}}{\sum_{h = 1}^K \text{e}^{\mathbf{w}_i^T \boldsymbol\delta_{h}}},~ k=1,\ldots,K,
\end{eqnarray}
where $\mathbf{w}_i$ is an $r\times 1$ vector of subject-level covariates, $\boldsymbol\delta_k$ is a $r\times 1$ vector of regression parameters associated with membership in cluster $k$. For identifiability purposes, we fix the reference category $k = K$ and set $\boldsymbol\delta_K = \mathbf{0}$. Under model (\ref{eq:pik}), $z_i|\boldsymbol\pi_i \sim \text{Multinomial}(1,\boldsymbol\pi_i)$, where $\boldsymbol\pi_i = (\pi_{1i},...,\pi_{Ki})$. As detailed below in Section \ref{s:postinf}, each cluster label $z_i$ is updated during MCMC estimation from its multinomial full conditional distribution and used in the remaining MCMC steps as the working cluster assignment for subject $i$. By allowing the cluster probabilities to vary across subjects, model (\ref{eq:mixture}) can be viewed as a \textit{mixture of experts} model, in which $\pi_{ki}$ acts as a \textit{gating function} controlling the prior probability of membership in cluster $k$, and $f(\mathbf{y}_i|\boldsymbol\theta_k)$ in equation (\ref{eq:mixture}) is the ``expert'' providing information on the within-cluster distribution of $\mathbf{y}_i$ (Bishop 2006).

To facilitate sampling, we adopt the efficient data-augmentation approach introduced by Polson {et al.} (2013), which expresses the inverse-logit function as a mixture P\'olya--Gamma densities. By using P\'olya--Gamma data augmentation for the multinomial model, we obtain a \textit{P\'olya--Gamma mixture of experts model} -- a computationally efficient way to obtain inferences for the mixing weights in the Bayesian setting. A random variable ${w}$ is said to follow a P\'olya--Gamma distribution with parameters $b > 0$ and $c \in \mathbb{R}$ if
\begin{equation}
{w} \sim \text{PG}(b,c) \stackrel{d}{=} \frac{1}{2\pi^2}\sum_{s=1}^{\infty}\frac{g_s}{(s-1/2)^2 + c^2/(4\pi^2)}, \label{eq:pg1}
\end{equation}
where $g_s \stackrel{iid}{\sim} \text{Ga}(b,1)$ for $s = 1,...,\infty$ . Polson {et al.} establish that for $a,\eta \in \mathbb{R}$,

@@

Polson et al. use these results to show that, for the logistic model, the Bernoulli likelihood can be written as a scale-mixture of normal densities with Polya-Gamma precision terms, $w$, resulting in a normal full conditional distribution for the logistic regression parameters. To extend the P\'olya--Gamma data augmentation approach to the multinomial setting, we first introduce the binary indicators $U_{ki}$, such that $U_{ki} = \mathds{1}_{(z_i = k)}$ -- i.e., $U_{ki}$ is an indicator variable taking the value $1$ if subject $i$ belongs to cluster $k$, and $0$ otherwise. The full conditional distribution of $\boldsymbol\delta_k$ is then given as  $$p(\boldsymbol\delta_k|\mathbf{U}_k,\boldsymbol\delta_{h \ne k}) \propto p(\boldsymbol\delta_k) \prod_{i = 1}^{n} \pi_{ki}^{U_{ki}}(1-\pi_{ki})^{1-U_{ki}},$$

@@

which is a logistic likelihood. We can thus apply the P\'olya--Gamma sampler described by Polson et al. for logistic regression to update each $\boldsymbol\delta_k$ one at a time based on the binary indicators $U_{ki}$. To do so, we first define for $k = 1,...,K$, the $n \times 1$ vector $\mathbf{U}^*_{k} = \left( \frac{U_{k1}-1/2}{w_{k1}},...,\frac{U_{kn}-1/2}{w_{kn}} \right )^T$. Polson et al. show that, conditional on $\boldsymbol{w} = (w_{k1},...,w_{kn})^T$, $\mathbf{U}^*_{k}$ follows a $\text{N}_n (\boldsymbol\eta_k,\mathbf{O}_k^{-1})$ distribution with mean $\boldsymbol\eta_k = (\eta_{k1},...,\eta_{kn})^T$ and precision matrix $\mathbf{O}_k = \text{Diag}(w_{k1},...,w_{kn})$. Thus, it follows that the full conditional distribution of $\boldsymbol\delta_k$ is given by 

@@

To accommodate intermittent missing at random (MAR) responses, we propose a convenient imputation algorithm that can be implemented ``online'' -- that is, as part of the Gibbs sampler. In Section 6, we discuss extensions to allow for non-ignorable missingness (i.e., observations missing not at random). Suppose $\mathbf{y}_i$ has $q_i\in (1,\ldots,J)$ observed values, denoted $\mathbf{y}^{obs}_i$, and $J-q_i$ intermittent missing values, denoted $\mathbf{y}^{miss}_i$. We can make use of the stochastic representation given in equation (\ref{eq:msnreg}) to impute $\mathbf{y}^{miss}_i$ from its conditional multivariate normal distribution given $(z_i,t_i,\mathbf{y}^{obs}_i)$:

@@

The location vector $\boldsymbol\mu_{ki}$ is defined as $\boldsymbol\mu_{ki} = \mathbf{X}_i \boldsymbol\beta_k + t_i \boldsymbol\psi_k$, and is partitioned into $\boldsymbol\mu_{ki}^{miss}$ and $\boldsymbol\mu_{ki}^{obs}$ with respect to the missing and observed indices of $\mathbf{y}_i$, respectively. The partition $\boldsymbol\Sigma_{k11}$ is a $(J-q_i) \times (J-q_i)$ matrix containing the rows and columns of $\boldsymbol\Sigma_k$ corresponding to $\mathbf{y}^{miss}_{i}$. Similarly, $\boldsymbol\Sigma_{k12}$ is a $(J-q_i) \times q_i$ matrix containing the rows of $\boldsymbol\Sigma_k$ that correspond to $\mathbf{y}^{miss}_{i}$, but columns of $\boldsymbol\Sigma_k$ that correspond to $\mathbf{y}^{obs}_{i}$. The remaining partitions $\boldsymbol\Sigma_{k21}, \ \text{and} \ \boldsymbol\Sigma_{k22}$ are defined in the same manner. These results follow from conventional multivariate normal theory. We note that, while conditional on $t_i$, $\mathbf{y}_i$ follows a MVN distribution, after marginalizing over $t_i$ the vector $\mathbf{y}_i$ follows a MSN distribution. Thus, this proposed online conditional imputation method provides a convenient way of imputing MSN responses using samples from more standard densities.

An attractive practical feature of this imputation algorithm is that it avoids multiplicative run-time scaling in $m$, the number of imputations (Gelman {et al.} 2013; Zhou and Reiter, 2010). Our approach also provides more opportunities to explore the missing data parameter space than does multiple imputation, since each missing component is drawn once per MCMC iteration, and often in practice $n_{sim} >> m$, where $n_{sim}$ is the total number of MCMC iterations (\textbf{find a reference}). In Section 4, we conduct simulation studies to demonstrate that imputing the missing MSN responses with our online conditional imputation method improves inferences over standard Bayesian multiple imputation as outlined by Gelman et al. (2013).

@@

\subsubsection{Prior Specification}
We adopt a fully Bayesian inferential approach and assign prior distributions to all model parameters. Conveniently, all parameters admit conditionally conjugate priors, which greatly improves posterior computation via a data-augmented Gibbs sampler. To make use of the matrix normal representation introduced previously in Section 3, we define the matrix of regression parameters $\mathbb{B}^*_k$, where $\text{vec}(\mathbf{B}^*_k) = \boldsymbol\beta^*_k = (\boldsymbol\beta_k^T,\boldsymbol\psi_k^T)^T$. We assign $\mathbf{B}^*_k|\boldsymbol\Sigma_k$ a $\text{MatNorm}(\mathbf{B}^*_{0k},\mathbf{I}_{p+1},\boldsymbol\Sigma_k)$ prior, where $\mathbf{B}^*_{0k}$ is a matrix of location parameters such that $\text{vec}(\mathbf{B}^*_{0k})$ is a vector of prior location parameters for the components of $\boldsymbol\beta^*_k$, $\mathbf{I}_{p+1}$ is the $(p+1)$-dimensional identity matrix, and $\boldsymbol\Sigma_k$ is the covariance matrix defined in equation (\ref{eq:msndens}), for which we specify an $\text{IW}(\mathbf{V}_{0k},\nu_{0k})$ prior. This leads to a matrix-normal-inverse-Wishart joint prior for $\mathbb{B}^*_k$ and $\boldsymbol\Sigma_k$, which is the conjugate joint prior for the regression parameters in the matrix normal model given in equation (\ref{eq:matsn}). This conjugate prior specification induces convenient closed-form full conditional distributions that can be easily updated within our proposed Gibbs sampler.

For the multinomial logit model, the regression parameters $\boldsymbol\delta_k = (\delta_{k1},...,\delta_{kr})^T$ are assigned a $\text{N}_r(\mathbf{d}_{0k}, \mathbf{S}_{0k})$ prior for $k = 1,...,K-1$, which is conditionally conjugate under the P\'olya-Gamma sampling scheme described in Section \ref{s:multinom}. We allow the normal-inverse-Wishart and multinomial hyperparameters (e.g., $\mathbf{B}^*_{0k}$ and $\mathbf{V}_{0k}$) to vary across clusters, though they may be shared across clusters in practice. An advantage of allowing for cluster-specific prior parameters is that \textit{a priori} knowledge of motor development trends can be incorporated into certain clusters while allowing the priors for other clusters to be less informative. Additionally, prior information regarding the effect of certain covariates on cluster membership can be incorporated in to the model by choosing informative values for $\mathbf{d}_{0k} \ \text{and} \ \mathbf{S}_{0k}$

\subsubsection{Posterior Inference} 
\label{s:postinf}
The above prior specification induces closed-form full conditionals for all model parameters, which can be efficiently updated as part of the Gibbs sampler outlined below. Additional details on derivations of full conditionals can be found in the Web Appendix A. We report MCMC diagnostics in Sections \ref{s:sim} and \ref{s:app}.

\paragraph{\textbf{Step 1}: Conditional MSN Imputation} The sampler begins by imputing missing values $\mathbf{y}_i^{miss}$ conditional on current values of $z_i = k$ and $t_i$ as well as the associated $\mathbf{y}_i^{obs}$ observed data vector. Specifically, for $i = 1,...,n$, we draw $\mathbf{y}_i^{miss}$ from $\text{N}_{J-q_i}(\boldsymbol\mu^{cond}_{ki},\boldsymbol\Sigma^{cond}_k)$ as described in equation (\ref{eq:impute}). We conclude by constructing a complete outcome vector $\mathbf{y}_i$ that merges $\mathbf{y}_i^{miss}$ with $\mathbf{y}_i^{obs}$.
